---
title: "Analysis for consensusDE paper - airway RNA-seq"
author: "Ashley J. Waardenberg"
date: "30/10/2019"
output:
  github_document:
    toc: yes
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{airway_results}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::github_document}
---

# Notes
30-10-2019
+ modifications in response to reviewers comments
+ updated to consensusDE version 1.3.4
+ addition of bland-altman analyses

30-06-2019
+ updated for compatability with consensusDE version 1.3.3
+ setting of norm_method = "all_defaults" for multi_de_pairs

# libraries used for analysis

```{r eval=TRUE, message=FALSE, warning=FALSE, include=TRUE}
library(consensusDE)
library(airway)
library(lattice)
library(blandr)
data(airway)
```

# functions used for analysis
```{r sourceCode,echo=TRUE}
source("./functions/return_airway_stats.r")
```

# Analysis of RNA-seq data with consensusDE
```{r eval=TRUE, message=FALSE, warning=FALSE, include=TRUE}
# establishment of data for pair-wise analysis
colData(airway)$group <- colData(airway)$dex
colData(airway)$file <- rownames(colData(airway))

# read into a summarized experiment, with low counts filtered
airway_filter <- buildSummarized(summarized = airway,
                                 filter = TRUE)

# run DE analysis with RUV
airway_ruv <- multi_de_pairs(summarized = airway_filter,
                             norm_method = "all_defaults",
                             ruv_correct = TRUE)

# run DE analysis without RUV
airway_NOruv <- multi_de_pairs(summarized = airway_filter,
                               norm_method = "all_defaults",
                               ruv_correct = FALSE)

```

# Jaccard Coefficient and Set Sizes

```{r}
# extract cde statistics
non_ruv <- return_cde_stats(airway_NOruv)
ruv <- return_cde_stats(airway_ruv)

# Jaccard coefficients
JC_stats <- rbind(unlist(non_ruv$props),
                  unlist(ruv$props)
                  )
rownames(JC_stats) <- c("non_RUV", "RUV")

#print results
JC_stats

# set sizes
non_ruv_set_size <- c(length(non_ruv$venn_input$VM),
                      length(non_ruv$venn_input$DE),
                      length(non_ruv$venn_input$ER),
                      length(non_ruv$intersect))

ruv_set_size <- c(length(ruv$venn_input$VM),
                  length(ruv$venn_input$DE),
                  length(ruv$venn_input$ER),
                  length(ruv$intersect))

# percetage increases (nonRUV to RUV)
percent_diff <- c((length(ruv$venn_input$VM)-length(non_ruv$venn_input$VM))/length(ruv$venn_input$VM),
                  (length(ruv$venn_input$DE)-length(non_ruv$venn_input$DE))/length(ruv$venn_input$DE),
                  (length(ruv$venn_input$ER)-length(non_ruv$venn_input$ER))/length(ruv$venn_input$ER),
                  (length(ruv$intersect)-length(non_ruv$intersect))/length(ruv$intersect))

set_sizes <- rbind(non_ruv_set_size,
                   ruv_set_size,
                   percent_diff)

colnames(set_sizes) <- c("voom", "DESeq2", "edgeR", "intersect")
rownames(set_sizes) <- c("non_RUV", "RUV", "percent_diff")

# print results
set_sizes
```

# Mean LogFC standard deviation of sets

```{r}

mean_sd <- rbind(c(mean(abs(non_ruv$ER$LogFC_sd)),
                   mean(abs(non_ruv$VM$LogFC_sd)),
                   mean(abs(non_ruv$DE$LogFC_sd))),
                 c(mean(abs(ruv$ER$LogFC_sd)),
                   mean(abs(ruv$VM$LogFC_sd)),
                   mean(abs(ruv$DE$LogFC_sd))))

colnames(mean_sd) <- c("edgeR", "voom", "DESeq2")
rownames(mean_sd) <- c("non_RUV", "RUV")

# print results
mean_sd

```

```{r}
# of the non-intersecting sets for each algorithm
mean_sd_not_intersect <- rbind(c(mean(abs(non_ruv$ER[!non_ruv$ER$ID %in% non_ruv$intersect,]$LogFC_sd)),
                                 mean(abs(non_ruv$VM[!non_ruv$VM$ID %in% non_ruv$intersect,]$LogFC_sd)),
                                 mean(abs(non_ruv$DE[!non_ruv$DE$ID %in% non_ruv$intersect,]$LogFC_sd))),
                               c(mean(abs(ruv$ER[!ruv$ER$ID %in% ruv$intersect,]$LogFC_sd)),
                                 mean(abs(ruv$VM[!ruv$VM$ID %in% ruv$intersect,]$LogFC_sd)),
                                 mean(abs(ruv$DE[!ruv$DE$ID %in% ruv$intersect,]$LogFC_sd))))

colnames(mean_sd_not_intersect) <- c("edgeR", "voom", "DESeq2")
rownames(mean_sd_not_intersect) <- c("non_RUV", "RUV")

# print results
mean_sd_not_intersect
```

```{r}
# overall intersect
mean_sd_intersect <- c(mean(abs(non_ruv$DE[non_ruv$DE$ID %in% non_ruv$intersect,]$LogFC_sd)),
                       mean(abs(ruv$DE[ruv$DE$ID %in% ruv$intersect,]$LogFC_sd)))

names(mean_sd_intersect) <- c("non_RUV", "RUV")

# print results
mean_sd_intersect
```

```{r}
# overall union
mean_sd_union <- c(mean(abs(unique(rbind(non_ruv$ER,non_ruv$VM,non_ruv$DE)$LogFC_sd))),
                   mean(abs(unique(rbind(ruv$ER,ruv$VM,ruv$DE)$LogFC_sd))))
names(mean_sd_union) <- c("non_RUV", "RUV")

# print results
mean_sd_union
```

# LogFC goodness of fit - extract statistics

```{r}
# extract statistics
all_FC <- get_merged_logfc(airway_NOruv)
all_FC_ruv <- get_merged_logfc(airway_ruv)

```

# LogFC goodness of fit - without RUV

```{r}
# extract logFC values for edgeR, voom and DESeq2
# using R-squared for comparison

# 1. Voom vs. DESeq2
plot(all_FC$logFC_VM, 
     all_FC$logFC_DE, 
     xlab="voom - logFC", 
     ylab="DESeq2 - logFC")
summary(lm(all_FC$logFC_VM~all_FC$logFC_DE))

# 2. Voom vs. edgeR
plot(all_FC$logFC_VM, 
     all_FC$logFC_ER,
     xlab="voom - logFC", 
     ylab="edgeR - logFC")
summary(lm(all_FC$logFC_VM~all_FC$logFC_ER))

# 3. edgeR vs. DESeq2
plot(all_FC$logFC_ER, 
     all_FC$logFC_DE, 
     xlab="edgeR - logFC", 
     ylab="DESeq2 - logFC")
summary(lm(all_FC$logFC_ER~all_FC$logFC_DE))

```

# LogFC goodness of fit - with RUV

```{r}
# extract logFC values for edgeR, voom and DESeq2
# using R-squared for comparison

# 1. Voom vs. DESeq2
plot(all_FC_ruv$logFC_VM, 
     all_FC_ruv$logFC_DE, 
     xlab="voom - logFC", 
     ylab="DESeq2 - logFC")
summary(lm(all_FC_ruv$logFC_VM~all_FC_ruv$logFC_DE))

# 2. Voom vs. edgeR
plot(all_FC_ruv$logFC_VM, 
     all_FC_ruv$logFC_ER,
     xlab="voom - logFC", 
     ylab="edgeR - logFC")
summary(lm(all_FC_ruv$logFC_VM~all_FC_ruv$logFC_ER))

# 3. edgeR vs. DESeq2
plot(all_FC_ruv$logFC_ER, 
     all_FC_ruv$logFC_DE, 
     xlab="edgeR - logFC", 
     ylab="DESeq2 - logFC")
summary(lm(all_FC_ruv$logFC_ER~all_FC_ruv$logFC_DE))

```

# Blant Altman analysis

```{r}
# bias of mean differences (or level of mean agreement)

all_FC_BA <- data.frame("v_de" = c(blandr.statistics(all_FC$logFC_VM , all_FC$logFC_DE, sig.level=0.95)$bias,
                                   blandr.statistics(all_FC_ruv$logFC_VM , all_FC_ruv$logFC_DE, sig.level=0.95)$bias),
                        "v_er" = c(blandr.statistics(all_FC$logFC_VM , all_FC$logFC_ER, sig.level=0.95)$bias,
                                   blandr.statistics(all_FC_ruv$logFC_VM , all_FC_ruv$logFC_ER, sig.level=0.95)$bias),
                        "de_er" = c(blandr.statistics(all_FC$logFC_DE , all_FC$logFC_ER, sig.level=0.95)$bias,
                                    blandr.statistics(all_FC_ruv$logFC_DE , all_FC_ruv$logFC_ER, sig.level=0.95)$bias))

row.names(all_FC_BA) <- c("nonRUV", "RUV")

#print results
all_FC_BA
```

```{r}
# limits of agreement (boundaries of agreement values between the different methods @ 95%)
# upperLOA and lowerLOA

FC_BA_upperLOA <- data.frame("v_de" = c(blandr.statistics(all_FC$logFC_VM , all_FC$logFC_DE, sig.level=0.95)$upperLOA,
                                       blandr.statistics(all_FC_ruv$logFC_VM , all_FC_ruv$logFC_DE, sig.level=0.95)$upperLOA),
                            "v_er" = c(blandr.statistics(all_FC$logFC_VM , all_FC$logFC_ER, sig.level=0.95)$upperLOA,
                                       blandr.statistics(all_FC_ruv$logFC_VM , all_FC_ruv$logFC_ER, sig.level=0.95)$upperLOA),
                            "de_er" = c(blandr.statistics(all_FC$logFC_DE , all_FC$logFC_ER, sig.level=0.95)$upperLOA,
                                        blandr.statistics(all_FC_ruv$logFC_DE , all_FC_ruv$logFC_ER, sig.level=0.95)$upperLOA)
                            )

row.names(FC_BA_upperLOA) <- c("nonRUV", "RUV")
#print results
FC_BA_upperLOA

```

```{r}
# barchart of results
FC_BA_upperLOA_bc <- data.frame("group" = names(unlist(FC_BA_upperLOA)),
                                "LOA" = unlist(FC_BA_upperLOA),
                                "reps" = rep(c("nonRUV", "RUV"),3))

barchart(LOA~group,
         data = FC_BA_upperLOA_bc,
         groups = reps,
         main = "upperLOA",
         auto.key = TRUE)
```




```{r}
FC_BA_lowerLOA <- data.frame("v_de" = c(blandr.statistics(all_FC$logFC_VM , all_FC$logFC_DE, sig.level=0.95)$lowerLOA,
                                       blandr.statistics(all_FC_ruv$logFC_VM , all_FC_ruv$logFC_DE, sig.level=0.95)$lowerLOA),
                            "v_er" = c(blandr.statistics(all_FC$logFC_VM , all_FC$logFC_ER, sig.level=0.95)$lowerLOA,
                                       blandr.statistics(all_FC_ruv$logFC_VM , all_FC_ruv$logFC_ER, sig.level=0.95)$lowerLOA),
                            "de_er" = c(blandr.statistics(all_FC$logFC_DE , all_FC$logFC_ER, sig.level=0.95)$lowerLOA,
                                        blandr.statistics(all_FC_ruv$logFC_DE , all_FC_ruv$logFC_ER, sig.level=0.95)$lowerLOA)
                            )

row.names(FC_BA_lowerLOA) <- c("nonRUV", "RUV")
#print results
FC_BA_lowerLOA

```


```{r}
# barchart of results
FC_BA_lowerLOA_bc <- data.frame("group" = names(unlist(FC_BA_lowerLOA)),
                                "LOA" = abs(unlist(FC_BA_lowerLOA)),
                                "reps" = rep(c("nonRUV", "RUV"),3))

barchart(LOA~group,
         data = FC_BA_lowerLOA_bc,
         groups = reps,
         main = "lowerLOA",
         auto.key = TRUE)


```

#  Blant - Altman plots a

```{r}
# plot results and determine statistics

# 1. Voom vs. DESeq2
blandr.draw(method1 = all_FC$logFC_VM, 
            method2 = all_FC$logFC_DE,
            plotTitle = "Bland-Altman, voom vs. DEseq2",
            annotate = TRUE,
            lowest_y_axis = -2,
            highest_y_axis = 2,
            plotter = "rplot")

# 2. Voom vs. edgeR
blandr.draw(method1 = all_FC$logFC_VM, 
            method2 = all_FC$logFC_ER,
            plotTitle = "Bland-Altman, voom vs. edgeR",
            annotate = TRUE,
            lowest_y_axis = -2,
            highest_y_axis = 2,
            plotter = "rplot")

# 3. edgeR vs. DESeq2
blandr.draw(method1 = all_FC$logFC_ER, 
            method2 = all_FC$logFC_DE,
            plotTitle = "Bland-Altman, edgeR vs. DEseq2",
            annotate = TRUE,
            lowest_y_axis = -2,
            highest_y_axis = 2,
            plotter = "rplot")

```

# add Blant - Altman plots and t-tests - with RUV 

```{r}

# 1. Voom vs. DESeq2
blandr.draw(method1 = all_FC_ruv$logFC_VM, 
            method2 = all_FC_ruv$logFC_DE,
            plotTitle = "Bland-Altman, voom vs. DEseq2",
            annotate = TRUE,
            lowest_y_axis = -2,
            highest_y_axis = 2,
            plotter = "rplot")

# 2. Voom vs. edgeR
blandr.draw(method1 = all_FC_ruv$logFC_VM, 
            method2 = all_FC_ruv$logFC_ER,
            plotTitle = "Bland-Altman, voom vs. edgeR",
            annotate = TRUE,
            lowest_y_axis = -2,
            highest_y_axis = 2,
            plotter = "rplot")

# 3. edgeR vs. DESeq2
blandr.draw(method1 = all_FC_ruv$logFC_ER, 
            method2 = all_FC_ruv$logFC_DE,
            plotTitle = "Bland-Altman, edgeR vs. DEseq2",
            annotate = TRUE,
            lowest_y_axis = -2,
            highest_y_axis = 2,
            plotter = "rplot")

```

# sessionInfo
```{r}
sessionInfo()
```

# functions used
printing for static reference

```{r}
return_cde_stats

get_merged_logfc
```
