---
title: "Analysis for consensusDE paper - simulation"
author: "Ashley J. Waardenberg"
date: "30/10/2019"
output:
  github_document:
    toc: yes
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{simulation_results}
  %\VignetteEngine{knitr::github_document}
  %\VignetteEncoding{UTF-8}
---

# Notes
30-10-2019
+ modifications in response to reviewers comments
+ updated to consensusDE version 1.3.4
+ addition of bland-altman analyses

30-06-2019
+ updated for compatability with consensusDE version 1.3.3
+ setting of norm_method = "all_defaults" for multi_de_pairs

# functions used for analysis

```{r sourceCode,echo=TRUE}
source("./functions/run_sims.r")
source("./functions/return_sims_stats.r")

# path for saving files
# where to download file, home directory, or change:
download_dir <- "~/"

```

# load libraries
```{r eval=TRUE, message=FALSE, warning=FALSE, include=TRUE}
library(consensusDE)
library(lattice)
library(metaseqR)
library(edgeR)
library(blandr)

```

# dataset used
```{r eval=TRUE, include=TRUE}

# Mouse - bottomly	2 inbred mouse strains
# phenodata
# http://bowtie-bio.sourceforge.net/recount/phenotypeTables/bottomly_phenodata.txt
# count table
bottomly <- "http://bowtie-bio.sourceforge.net/recount/countTables/bottomly_count_table.txt"

download.file(url = bottomly,
              destfile = paste(download_dir, "bottomly_count_table.txt", sep=""))

```

# run simulation

parameters
- p_thresh = threshold for determining significance.
- norm_method = "all_deafults". Selects the default normalisation approach for 
each method.

```{r eval=TRUE, message=FALSE, warning=FALSE, include=TRUE}
# establish datasets to model distribution
data_to_test <- c(paste(download_dir, "bottomly_count_table.txt", sep=""))

# number of replicates to test
reps <- c(3, 5)

# default parameters for simulations:
# sim_number = 10
# de_number = 500
# de_size = 10000

# 1. with RUV
ruv <- lapply(seq_along(reps), function(x) 
                              wrap_all(data_to_test,
                                       p_thresh = 0.05,
                                       replicates = reps[x],
                                       ruv = TRUE,
                                       norm_method = "all_defaults"))
# 2. withOUT RUV
NOruv <- lapply(seq_along(reps), function(x) 
                              wrap_all(data_to_test,
                                       p_thresh = 0.05,
                                       replicates = reps[x],
                                       ruv = FALSE,
                                       norm_method = "all_defaults"))

```

# obtain mean R-squared of logFC

```{r}
# non-ruv
NOruv_3rep <- get_mean_FC(NOruv, 1) # 1 indicates experiment with 3 replicates
NOruv_5rep <- get_mean_FC(NOruv, 2) # 2 indicates experiment with 5 replicates
# ruv
ruv_3rep <- get_mean_FC(ruv, 1)
ruv_5rep <- get_mean_FC(ruv, 2)

# put all results into a table
table_R <- data.frame(c(as.numeric(apply(NOruv_3rep, 2, mean)), 
                        as.numeric(apply(ruv_3rep, 2, mean)),
                        as.numeric(apply(NOruv_5rep, 2, mean)),
                        as.numeric(apply(ruv_5rep, 2, mean))),
                      c(rep(c("3-NOruv", "3-ruv"), each = 3, 1), 
                        rep(c("5-NOruv", "5-ruv"), each = 3, 1)),
                      c(rep(c("1-VM.DE", "2-VM.ER", "3-ER.DE"), 4)))
colnames(table_R) <- c("R", "reps", "group")

# barchart of results
barchart(R~group,
         data = table_R,
         groups = reps,
         main = "R-squared of logFC with or without RUV",
         auto.key = TRUE,
         ylim = c(0, 1.1))

```

# Bland-Altman analysis
```{r}
# non-ruv
NOruv_3rep <- get_bias_FC(NOruv, 1, method = "bias") # 1 indicates experiment with 3 replicates
NOruv_5rep <- get_bias_FC(NOruv, 2, method = "bias") # 2 indicates experiment with 5 replicates

# ruv
ruv_3rep <- get_bias_FC(ruv, 1, method = "bias")
ruv_5rep <- get_bias_FC(ruv, 2, method = "bias")

# put all results into a table
# for bias use the absolute values.
table_bias <- data.frame(c(as.numeric(apply(abs(NOruv_3rep), 2, mean)), 
                        as.numeric(apply(abs(ruv_3rep), 2, mean)),
                        as.numeric(apply(abs(NOruv_5rep), 2, mean)),
                        as.numeric(apply(abs(ruv_5rep), 2, mean))),
                      c(rep(c("3-NOruv", "3-ruv"), each = 3, 1), 
                        rep(c("5-NOruv", "5-ruv"), each = 3, 1)),
                      c(rep(c("1-VM.DE", "2-VM.ER", "3-ER.DE"), 4)))
colnames(table_bias) <- c("bias", "reps", "group")

# barchart of results
barchart(bias~group,
         data = table_bias,
         groups = reps,
         main = "Bias of logFC with or without RUV",
         auto.key = TRUE)

```

# upper limits
```{r}
# non-ruv
NOruv_3rep <- get_bias_FC(NOruv, 1, method = "upperLOA") # 1 indicates experiment with 3 replicates
NOruv_5rep <- get_bias_FC(NOruv, 2, method = "upperLOA") # 2 indicates experiment with 5 replicates

# ruv
ruv_3rep <- get_bias_FC(ruv, 1, method = "upperLOA")
ruv_5rep <- get_bias_FC(ruv, 2, method = "upperLOA")

# put all results into a table
# for bias use the absolute values.
table_upperLOA <- data.frame(c(as.numeric(apply(abs(NOruv_3rep), 2, mean)), 
                        as.numeric(apply(abs(ruv_3rep), 2, mean)),
                        as.numeric(apply(abs(NOruv_5rep), 2, mean)),
                        as.numeric(apply(abs(ruv_5rep), 2, mean))),
                      c(rep(c("3-NOruv", "3-ruv"), each = 3, 1), 
                        rep(c("5-NOruv", "5-ruv"), each = 3, 1)),
                      c(rep(c("1-VM.DE", "2-VM.ER", "3-ER.DE"), 4)))
colnames(table_upperLOA) <- c("LOA", "reps", "group")

# barchart of results
barchart(LOA~group,
         data = table_upperLOA,
         groups = reps,
         main = "table_upperLOA of logFC with or without RUV",
         ylim = c(0,0.2),
         auto.key = TRUE)
```

# lower limits
```{r}
# non-ruv
NOruv_3rep <- get_bias_FC(NOruv, 1, method = "lowerLOA") # 1 indicates experiment with 3 replicates
NOruv_5rep <- get_bias_FC(NOruv, 2, method = "lowerLOA") # 2 indicates experiment with 5 replicates

# ruv
ruv_3rep <- get_bias_FC(ruv, 1, method = "lowerLOA")
ruv_5rep <- get_bias_FC(ruv, 2, method = "lowerLOA")

# put all results into a table
# for bias use the absolute values.
table_lowerLOA <- data.frame(c(as.numeric(apply(abs(NOruv_3rep), 2, mean)), 
                        as.numeric(apply(abs(ruv_3rep), 2, mean)),
                        as.numeric(apply(abs(NOruv_5rep), 2, mean)),
                        as.numeric(apply(abs(ruv_5rep), 2, mean))),
                      c(rep(c("3-NOruv", "3-ruv"), each = 3, 1), 
                        rep(c("5-NOruv", "5-ruv"), each = 3, 1)),
                      c(rep(c("1-VM.DE", "2-VM.ER", "3-ER.DE"), 4)))
colnames(table_lowerLOA) <- c("LOA", "reps", "group")

# barchart of results
barchart(LOA~group,
         data = table_lowerLOA,
         groups = reps,
         main = "lowerLOA of logFC with or without RUV",
         ylim = c(0,0.2),
         auto.key = TRUE)
```

# Fold change SD

```{r}
# Fold Change Statistics
# non-ruv
NOruv_3rep_stats_fc <- clean_stats(NOruv, 1, isLogFC = TRUE)
NOruv_5rep_stats_fc <- clean_stats(NOruv, 2, isLogFC = TRUE)
# ruv
ruv_3rep_stats_fc <- clean_stats(ruv, 1, isLogFC = TRUE)
ruv_5rep_stats_fc <- clean_stats(ruv, 2, isLogFC = TRUE)

# Obtain the mean Fold-change Standard Deviations (SD)
# without RUV - 3 reps
mean_3rep_fc <- sapply(1:ncol(NOruv_3rep_stats_fc), function(x) mean(unlist(data.frame(NOruv_3rep_stats_fc)[x]), na.rm=TRUE))
names(mean_3rep_fc) <- colnames(NOruv_3rep_stats_fc)
# without RUV - 5 reps
mean_5rep_fc <- sapply(1:ncol(NOruv_5rep_stats_fc), function(x) mean(unlist(data.frame(NOruv_5rep_stats_fc)[x]), na.rm=TRUE))
names(mean_5rep_fc) <- colnames(NOruv_5rep_stats_fc)

# with RUV - 3 reps
mean_3rep_fc_ruv <- sapply(1:ncol(ruv_3rep_stats_fc), function(x) mean(unlist(data.frame(ruv_3rep_stats_fc)[x]), na.rm=TRUE))
names(mean_3rep_fc_ruv) <- colnames(ruv_3rep_stats_fc)
# with RUV - 5 reps
mean_5rep_fc_ruv <- sapply(1:ncol(ruv_5rep_stats_fc), function(x) mean(unlist(data.frame(ruv_5rep_stats_fc)[x]), na.rm=TRUE))
names(mean_5rep_fc_ruv) <- colnames(ruv_5rep_stats_fc)

# summary of results
mean_fc_summary <- rbind(mean_3rep_fc,
                         mean_5rep_fc,
                         mean_3rep_fc_ruv,
                         mean_5rep_fc_ruv)

# print results
mean_fc_summary

```

# Jaccard Coeffecient
```{r}
# non-ruv
NOruv_3rep_stats <- clean_stats(NOruv, 1)
NOruv_5rep_stats <- clean_stats(NOruv, 2)
# ruv
ruv_3rep_stats <- clean_stats(ruv, 1)
ruv_5rep_stats <- clean_stats(ruv, 2)

# JC for each set
# here, the first set is the intersect of all values, will normalise to 1
# 3 reps
int_3rep <- apply(NOruv_3rep_stats$TP_n, 2, mean) + apply(NOruv_3rep_stats$FP_n, 2, mean)
JC_3rep <- int_3rep[1]/int_3rep
# 5 reps
int_5rep <- apply(NOruv_5rep_stats$TP_n, 2, mean) + apply(NOruv_5rep_stats$FP_n, 2, mean)
JC_5rep <- int_5rep[1]/int_5rep

# 3 reps - with RUV
int_3rep_ruv <- apply(ruv_3rep_stats$TP_n, 2, mean) + apply(ruv_3rep_stats$FP_n, 2, mean)
JC_3rep_ruv <- int_3rep_ruv[1]/int_3rep_ruv
# 5 reps - with RUV
int_5rep_ruv <- apply(ruv_5rep_stats$TP_n, 2, mean) + apply(ruv_5rep_stats$FP_n, 2, mean)
JC_5rep_ruv <- int_5rep_ruv[1]/int_5rep_ruv

# summarize results into table
table_JC <- data.frame(c(as.numeric(JC_3rep[2:length(JC_3rep)]), 
                    as.numeric(JC_3rep_ruv[2:length(JC_3rep_ruv)]),
                    as.numeric(JC_5rep[2:length(JC_5rep)]), 
                    as.numeric(JC_5rep_ruv[2:length(JC_5rep_ruv)])),
                  c(rep(c("3-NOruv", "3-ruv"), each = 4, 1), 
                    rep(c("5-NOruv", "5-ruv"), each = 4, 1)),
                  c(rep(c("5-union", "4-EdgeR", "2-voom", "3-DESeq2"), 4)))
colnames(table_JC) <- c("JC", "reps", "group")

# plot results
barchart(JC~group,
         data = table_JC,
         groups = reps,
         main = "Jaccard Coefficient, relative to total intersect",
         auto.key = TRUE,
         ylim = c(0, 1.05))

```

# Set sizes for each comparison

```{r}
# set sizes (genes call DE) for each method
size_summary <- rbind(int_3rep,
                      int_3rep_ruv,
                      int_5rep,
                      int_5rep_ruv)

size_summary
```


# FDR
```{r}
# put FDR (mean from all simulations) into table
table_FDR <- data.frame(c(as.numeric(apply(NOruv_3rep_stats$FDR, 2, mean)), 
                          as.numeric(apply(ruv_3rep_stats$FDR, 2, mean)),
                          as.numeric(apply(NOruv_5rep_stats$FDR, 2, mean)),
                          as.numeric(apply(ruv_5rep_stats$FDR, 2, mean))),
                        c(rep(c("3-NOruv", "3-ruv"), each = 5, 1), 
                          rep(c("5-NOruv", "5-ruv"), each = 5, 1)),
                        c(rep(c("1-intersect", "5-union", "4-EdgeR", "2-voom", "3-DESeq2"), 4)))
colnames(table_FDR) <- c("FDR", "reps", "group")

# plot FDR
barchart(FDR~group,
         data = table_FDR,
         groups = reps,
         main = "False Discovery Rates",
         auto.key = TRUE)

```

# sensitivity - added in revision (recall)
```{r}
# put sensitivity (mean from all simulations) into table
table_sens <- data.frame(c(as.numeric(apply(NOruv_3rep_stats$sens, 2, mean)), 
                          as.numeric(apply(ruv_3rep_stats$sens, 2, mean)),
                          as.numeric(apply(NOruv_5rep_stats$sens, 2, mean)),
                          as.numeric(apply(ruv_5rep_stats$sens, 2, mean))),
                        c(rep(c("3-NOruv", "3-ruv"), each = 5, 1), 
                          rep(c("5-NOruv", "5-ruv"), each = 5, 1)),
                        c(rep(c("1-intersect", "5-union", "4-EdgeR", "2-voom", "3-DESeq2"), 4)))
colnames(table_sens) <- c("sens", "reps", "group")

# plot
barchart(sens~group,
         data = table_sens,
         groups = reps,
         main = "Sensitivity",
         ylim = c(0, 1.05),
         auto.key = TRUE)
```

# F1 statistics

```{r}

table_F1 <- data.frame(c(as.numeric(apply(NOruv_3rep_stats$F1, 2, mean)), 
                    as.numeric(apply(ruv_3rep_stats$F1, 2, mean)),
                    as.numeric(apply(NOruv_5rep_stats$F1, 2, mean)),
                      as.numeric(apply(ruv_5rep_stats$F1, 2, mean))),
                  c(rep(c("3-NOruv", "3-ruv"), each = 5, 1), 
                    rep(c("5-NOruv", "5-ruv"), each = 5, 1)),
                  c(rep(c("1-intersect", "5-union", "4-EdgeR", "2-voom", "3-DESeq2"), 4)))
colnames(table_F1) <- c("F1", "reps", "group")

barchart(F1~group,
         data = table_F1,
         groups = reps,
         main = "F1 - statistics",
         auto.key = TRUE,
         ylim = c(0, 1.05))
         
```

# ACCURACY statistic - not good for imbalanced data

```{r}
table_ACC <- data.frame(c(as.numeric(apply(NOruv_3rep_stats$ACC, 2, mean)), 
                    as.numeric(apply(ruv_3rep_stats$ACC, 2, mean)),
                    as.numeric(apply(NOruv_5rep_stats$ACC, 2, mean)),
                      as.numeric(apply(ruv_5rep_stats$ACC, 2, mean))),
                  c(rep(c("3-NOruv", "3-ruv"), each = 5, 1), 
                    rep(c("5-NOruv", "5-ruv"), each = 5, 1)),
                  c(rep(c("1-intersect", "5-union", "4-EdgeR", "2-voom", "3-DESeq2"), 4)))
colnames(table_ACC) <- c("ACC", "reps", "group")


barchart(ACC~group,
         data = table_ACC,
         groups = reps,
         main = "Accuracy",
         ylim = c(0, 1.05),
         auto.key = TRUE)

```

# sessionInfo
```{r}
sessionInfo()
```

# functions used
printing for static reference

```{r}
wrap_all

run_sims

return_logFC_sd

return_stats

cont_list

cont

get_mean_FC

get_r

clean_stats


```
